---
title: "Census PEP Population Targets, 1900-2025"
author: "Joshua Born"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
library(censusapi)
library(dplyr)
library(here)
library(readr)
library(readxl)
library(tidyr)

census_key <- Sys.getenv("CENSUS_KEY")
if (identical(census_key, "")) {
  stop("Set CENSUS_KEY (e.g. in .Renviron) to a free Census API key: https://api.census.gov/data/key_signup.html")
}
Sys.setenv(CENSUS_KEY = census_key)

data_raw_dir <- here("data-raw")
dir.create(data_raw_dir, showWarnings = FALSE, recursive = TRUE)

artifacts_dir <- here("output")
dir.create(artifacts_dir, showWarnings = FALSE, recursive = TRUE)

knitr::opts_chunk$set(
  cache = TRUE,
  echo = TRUE,
  tidy = TRUE
)
```

```{r download-if-missing, dependson="setup"}
# Downloads `url` to `dest_path` only if `dest_path` doesn't already exist,
# so repeated renders don't re-fetch ~30 files from Census on every knit.
download_if_missing <- function(url, dest_path) {
  if (!file.exists(dest_path)) {
    dir.create(dirname(dest_path), showWarnings = FALSE, recursive = TRUE)
    download.file(url, dest_path, mode = "wb", quiet = TRUE)
  }
  dest_path
}
```

```{r age-group-levels, dependson="setup"}
# Standard 18-bin 5-year age scheme (Census's own AGEGROUP coding,
# confirmed against the 2000-2009 pep/int_charagegroups product and the
# 2010-2020 intercensal workbook's own row labels).
age_group_levels <- c(
  "Under 5", "5 to 9", "10 to 14", "15 to 19", "20 to 24", "25 to 29",
  "30 to 34", "35 to 39", "40 to 44", "45 to 49", "50 to 54", "55 to 59",
  "60 to 64", "65 to 69", "70 to 74", "75 to 79", "80 to 84",
  "85 years and over"
)
```

```{r bin-age-5yr, dependson=c("setup", "age-group-levels")}
# Bins single-year ages into age_group_levels. `top_bin_age` is the age
# value the source uses for its own open-ended top code (85 for every era
# except 1900-1939, whose source only goes up to "75 years and over").
bin_age_5yr <- function(age, top_bin_age = 85) {
  if (top_bin_age >= 85) {
    breaks <- c(seq(0, 85, by = 5), Inf)
    labels <- age_group_levels
  } else {
    breaks <- c(seq(0, top_bin_age, by = 5), Inf)
    labels <- c(
      age_group_levels[seq_len(top_bin_age / 5)],
      paste(top_bin_age, "years and over")
    )
  }
  as.character(cut(age, breaks = breaks, labels = labels, right = FALSE))
}
```

```{r check-allages-reconciliation, dependson="setup"}
# Consistency check: does a by-age table, summed across AgeGroup,
# reproduce an existing all-ages table exactly?
check_allages_reconciliation <- function(by_age, allages, group_vars) {
  by_age |>
    group_by(across(all_of(group_vars))) |>
    summarize(PopulationFromByAge = sum(Population), .groups = "drop") |>
    inner_join(allages, by = group_vars) |>
    mutate(Diff = PopulationFromByAge - Population) |>
    summarize(MaxAbsDiff = max(abs(Diff)), .groups = "drop")
}
```

```{r check-adults-reconciliation, dependson=c("setup", "age-group-levels")}
# Consistency check: do a by-age table's bins from "20 to 24" up, plus a
# separately supplied Population_1819 (exact 18-19 count, or the
# documented 2/5 "15 to 19" interpolation for 1990-1999), reproduce an
# existing exact ages-18+ table?
check_adults_reconciliation <- function(by_age, age_18_19, adults, group_vars) {
  young_bins <- age_group_levels[1:4]
  by_age |>
    filter(!AgeGroup %in% young_bins) |>
    group_by(across(all_of(group_vars))) |>
    summarize(Population_FullBins = sum(Population), .groups = "drop") |>
    full_join(age_18_19, by = group_vars) |>
    mutate(
      Population_FullBins = replace_na(Population_FullBins, 0),
      Population_1819 = replace_na(Population_1819, 0),
      PopulationFromByAge = Population_FullBins + Population_1819
    ) |>
    inner_join(adults, by = group_vars) |>
    mutate(Diff = PopulationFromByAge - Population) |>
    summarize(MaxAbsDiff = max(abs(Diff)), .groups = "drop")
}
```

```{r race-scheme-collapse-maps, dependson="setup"}
# Race->Race lookups for deriving a coarser RaceScheme from a finer one.
# Only aggregation (coarser from finer) is ever done; a coarser scheme is
# never split back into a finer one.
race4_to_race3 <- c(White = "White", Black = "Black", AIAN = "Other", API = "Other")
race4_to_race2 <- c(White = "White", Black = "Nonwhite", AIAN = "Nonwhite", API = "Nonwhite")
race6_to_race3 <- c(
  White = "White", Black = "Black", AIAN = "Other",
  Asian = "Other", NHPI = "Other", TwoOrMore = "Other"
)
race6_to_race2 <- c(
  White = "White", Black = "Nonwhite", AIAN = "Nonwhite",
  Asian = "Nonwhite", NHPI = "Nonwhite", TwoOrMore = "Nonwhite"
)
race3_to_race2 <- c(White = "White", Black = "Nonwhite", Other = "Nonwhite")

# Applies a Race->Race collapse map to a table whose value column is
# `Population`, tags the result with the coarser `scheme`, and re-sums
# `Population` within the resulting groups (every other column must already
# be a grouping variable).
collapse_race <- function(data, mapping, scheme) {
  data |>
    mutate(Race = mapping[Race], RaceScheme = scheme) |>
    group_by(across(-Population)) |>
    summarize(Population = sum(Population), .groups = "drop")
}

# Same as collapse_race(), for the "18-19 exact/interpolated split" tables,
# whose value column is `Population_1819` instead of `Population`.
collapse_race_1819 <- function(data, mapping, scheme) {
  data |>
    mutate(Race = mapping[Race], RaceScheme = scheme) |>
    group_by(across(-Population_1819)) |>
    summarize(Population_1819 = sum(Population_1819), .groups = "drop")
}
```

## Overview

Builds two long/tidy U.S. Census Bureau Population Estimates Program (PEP)
population-target CSVs for raking/post-stratification, 1900-2025, by Year,
RaceScheme, Race, Sex, AgeGroup, and (from 1980 on) HispanicOrigin:
`census_targets.csv` (all ages) and `census_targets_adults.csv`
(ages 18+). Every input is downloaded directly from Census (cached under
`data-raw/`, gitignored, so repeated renders don't re-fetch); every chunk is
knitr-cached and `dependson`-chained so an unchanged upstream re-knits fast.

The underlying sources use four genuinely different race category schemes,
not one, because Census's own race questions changed over time. Rather than
collapse everything down to the coarsest shared scheme, both files carry
every scheme a given year supports, tagged by a `RaceScheme` column:

| RaceScheme | Race levels | Years |
|---|---|---|
| `race2` | White, Nonwhite | 1900-2025 |
| `race3` | White, Black, Other | 1960-2025 |
| `race4` | White, Black, AIAN, API | 1980-1999 |
| `race6` | White, Black, AIAN, Asian, NHPI, TwoOrMore | 2000-2025 |

A user should filter to a single `RaceScheme` value before summing
`Population`, since the four schemes' `Race` values are not compatible with
each other. Coarser schemes are always derived by collapsing a year's finest
available scheme, never estimated independently, so `race2` and `race3`
rows agree with `race4`/`race6` rows by construction wherever both exist for
the same year. `race4` stops at 1999 rather than continuing to 2025, because
Census's 2000+ "Two or More Races" category has no principled home in the
1980s/1990s-defined four-category scheme.

`HispanicOrigin` is `NA` before 1980 (no source before then carries a
Hispanic-origin breakdown) and Hispanic/NonHispanic-crossed from 1980 on,
with no redundant NA row alongside the 1980+ crossed rows (so summing all
rows for a cell never double-counts). This crossing is populated for every
`RaceScheme`'s own 1980-or-later rows, including `race4`, not just the
coarser schemes. See the full derivation, source discussion, and validation
(discontinuity audit, trend plots) in the `borninquisitive` repo's
`EDA/20260831_PEP_Population_Targets_for_Raking.Rmd`.

## Data Sources by Era

### 1900-1959: pre-1960 national age/sex/race tables (White/Nonwhite)

Six decade-bundle workbooks, `pe-11-1900s.xls` through `pe-11-1950s.xls`, one
worksheet per year, from the Census Bureau's
["National Intercensal Tables: 1900-1990"](https://www.census.gov/data/tables/time-series/demo/popest/pre-1980-national.html)
page. Two structural breaks occur within this range (both stated in the
workbooks' own footnote text): the open top age bin widens from "75 years
and over" to "85 years and over" at 1939/1940, and geographic coverage
changes (excludes, then includes, Alaska/Hawaii) at 1949/1950 -- not
corrected for here.

This section expects all 60 years from 1900 through 1959. The years listed
below confirm every year was found with no gaps.

```{r extract-1900-1959-from-xls, dependson=c("setup", "download-if-missing")}
decade_files_2cat <- c(
  "pe-11-1900s.xls", "pe-11-1910s.xls", "pe-11-1920s.xls",
  "pe-11-1930s.xls", "pe-11-1940s.xls", "pe-11-1950s.xls"
)

for (f in decade_files_2cat) {
  download_if_missing(
    paste0("https://www2.census.gov/programs-surveys/popest/tables/1900-1980/national/asrh/", f),
    file.path(data_raw_dir, f)
  )
}

read_pe11_2cat_sheet <- function(file, year) {
  # 1939->1940 top-age-bin widening, noted above.
  top_bin <- if (year < 1940) "75+" else "85+"
  top_bin_age <- if (year < 1940) 75 else 85

  raw <- suppressMessages(read_excel(
    file.path(data_raw_dir, file),
    sheet = as.character(year),
    col_names = FALSE,
    col_types = "text"
  ))
  tibble(
    Year = year,
    AgeRaw = raw[[1]],
    `All races.Total` = raw[[2]],
    `All races.Male` = raw[[3]],
    `All races.Female` = raw[[4]],
    `White.Total` = raw[[5]],
    `White.Male` = raw[[6]],
    `White.Female` = raw[[7]],
    `Nonwhite.Total` = raw[[8]],
    `Nonwhite.Male` = raw[[9]],
    `Nonwhite.Female` = raw[[10]]
  ) |>
    filter(grepl("^[0-9]+$", AgeRaw) | AgeRaw == top_bin) |>
    mutate(
      Age = as.numeric(if_else(AgeRaw == top_bin, as.character(top_bin_age), AgeRaw)),
      across(`All races.Total`:`Nonwhite.Female`, as.numeric)
    ) |>
    select(Year, Age, `All races.Total`:`Nonwhite.Female`)
}

year_file_pairs_2cat <- bind_rows(lapply(decade_files_2cat, function(f) {
  tibble(file = f, year = as.integer(excel_sheets(file.path(data_raw_dir, f))))
}))

Census_1900_1959_by_age <- bind_rows(Map(
  read_pe11_2cat_sheet,
  year_file_pairs_2cat$file,
  year_file_pairs_2cat$year
))

cat("Years extracted:", paste(sort(unique(Census_1900_1959_by_age$Year)), collapse = ", "))
```

```{r aggregate-1900-1959, dependson="extract-1900-1959-from-xls"}
aggregate_2cat <- function(by_age_data) {
  by_age_data |>
    group_by(Year) |>
    summarize(across(`All races.Total`:`Nonwhite.Female`, sum), .groups = "drop")
}

reshape_2cat <- function(totals) {
  totals |>
    select(Year, starts_with("White."), starts_with("Nonwhite.")) |>
    pivot_longer(
      cols = -Year,
      names_to = c("Race", "Sex"),
      names_sep = "\\.",
      values_to = "Population"
    ) |>
    filter(Sex %in% c("Male", "Female"))
}

Census_1900_1959_totals_adults <- aggregate_2cat(Census_1900_1959_by_age |> filter(Age >= 18))
Census_1900_1959_totals_allages <- aggregate_2cat(Census_1900_1959_by_age)

Census_1900_1959_adults <- reshape_2cat(Census_1900_1959_totals_adults) |>
  mutate(RaceScheme = "race2")
Census_1900_1959_allages <- reshape_2cat(Census_1900_1959_totals_allages) |>
  mutate(RaceScheme = "race2")
```

This is the only era with no coarser or finer RaceScheme to derive: the
source reports White and Nonwhite and nothing else, so `race2` is both the
finest and the only scheme available for these years. The check below
verifies that Male and Female sum to the reported total and that White and
Nonwhite sum to the reported total, both of which should be exactly zero.

```{r check-1900-1959, dependson="aggregate-1900-1959"}
Census_1900_1959_totals_adults |>
  mutate(
    Diff.MaleFemale = `All races.Total` - `All races.Male` - `All races.Female`,
    Diff.RaceSum = `All races.Total` - `White.Total` - `Nonwhite.Total`
  ) |>
  summarize(MaxAbsDiff.MaleFemale = max(abs(Diff.MaleFemale)), MaxAbsDiff.RaceSum = max(abs(Diff.RaceSum)))
```

### 1960-1969: pre-1970 national age/sex/race table (White/Black/Other)

`pe-11-1960s.xls`, same landing page and layout as above, but already using
the White/Black/Other scheme that carries through 2025.

This section expects all 10 years from 1960 through 1969. The years listed
below confirm every year was found with no gaps.

```{r extract-1960s-from-xls, dependson=c("setup", "download-if-missing")}
invisible(download_if_missing(
  "https://www2.census.gov/programs-surveys/popest/tables/1900-1980/national/asrh/pe-11-1960s.xls",
  file.path(data_raw_dir, "pe-11-1960s.xls")
))

read_1960s_sheet <- function(year) {
  raw <- suppressMessages(read_excel(
    file.path(data_raw_dir, "pe-11-1960s.xls"),
    sheet = as.character(year),
    col_names = FALSE,
    col_types = "text"
  ))
  tibble(
    Year = year,
    AgeRaw = raw[[1]],
    `All races.Total` = raw[[2]],
    `All races.Male` = raw[[3]],
    `All races.Female` = raw[[4]],
    `White.Total` = raw[[5]],
    `White.Male` = raw[[6]],
    `White.Female` = raw[[7]],
    `Black.Total` = raw[[8]],
    `Black.Male` = raw[[9]],
    `Black.Female` = raw[[10]]
  ) |>
    filter(grepl("^[0-9]+$", AgeRaw) | AgeRaw == "85+") |>
    mutate(
      Age = as.numeric(if_else(AgeRaw == "85+", "85", AgeRaw)),
      across(`All races.Total`:`Black.Female`, as.numeric)
    ) |>
    select(Year, Age, `All races.Total`:`Black.Female`)
}

Census_1960s_by_age <- bind_rows(lapply(1960:1969, read_1960s_sheet))

cat("Years extracted:", paste(sort(unique(Census_1960s_by_age$Year)), collapse = ", "))
```

```{r aggregate-1960s, dependson=c("extract-1960s-from-xls", "race-scheme-collapse-maps")}
aggregate_3cat <- function(by_age_data) {
  by_age_data |>
    group_by(Year) |>
    summarize(across(`All races.Total`:`Black.Female`, sum), .groups = "drop") |>
    mutate(
      `Other.Total` = `All races.Total` - `White.Total` - `Black.Total`,
      `Other.Male` = `All races.Male` - `White.Male` - `Black.Male`,
      `Other.Female` = `All races.Female` - `White.Female` - `Black.Female`
    )
}

reshape_3cat <- function(totals) {
  totals |>
    select(Year, starts_with("White."), starts_with("Black."), starts_with("Other.")) |>
    pivot_longer(
      cols = -Year,
      names_to = c("Race", "Sex"),
      names_sep = "\\.",
      values_to = "Population"
    ) |>
    filter(Sex %in% c("Male", "Female"))
}

# Tags a race3 (White/Black/Other) table with RaceScheme and adds its
# race2-collapsed derivative, for eras whose native scheme is race3.
add_race3_and_race2 <- function(race3_table) {
  race3 <- race3_table |> mutate(RaceScheme = "race3")
  bind_rows(race3, collapse_race(race3, race3_to_race2, "race2"))
}

# Same as add_race3_and_race2(), for the "18-19 exact split" tables, whose
# value column is Population_1819 instead of Population.
add_race3_and_race2_1819 <- function(race3_table) {
  race3 <- race3_table |> mutate(RaceScheme = "race3")
  bind_rows(race3, collapse_race_1819(race3, race3_to_race2, "race2"))
}

Census_1960s_totals_adults <- aggregate_3cat(Census_1960s_by_age |> filter(Age >= 18))
Census_1960s_totals_allages <- aggregate_3cat(Census_1960s_by_age)

Census_1960s_adults <- add_race3_and_race2(reshape_3cat(Census_1960s_totals_adults))
Census_1960s_allages <- add_race3_and_race2(reshape_3cat(Census_1960s_totals_allages))
```

The check below verifies that Male and Female sum to the reported total and
that White, Black, and Other sum to the reported total, both of which
should be exactly zero.

```{r check-1960s, dependson="aggregate-1960s"}
Census_1960s_totals_adults |>
  mutate(
    Diff.MaleFemale = `All races.Total` - `All races.Male` - `All races.Female`,
    Diff.RaceSum = `All races.Total` - `White.Total` - `Black.Total` - `Other.Total`
  ) |>
  summarize(MaxAbsDiff.MaleFemale = max(abs(Diff.MaleFemale)), MaxAbsDiff.RaceSum = max(abs(Diff.RaceSum)))
```

### 1970-1979: pre-1980 national age/sex/race table

`pe-11-1970s.xls`, same source and layout as 1960-1969. This section
expects all 10 years from 1970 through 1979. The years listed below
confirm every year was found with no gaps.

```{r extract-1970s-from-xls, dependson=c("setup", "download-if-missing")}
invisible(download_if_missing(
  "https://www2.census.gov/programs-surveys/popest/tables/1900-1980/national/asrh/pe-11-1970s.xls",
  file.path(data_raw_dir, "pe-11-1970s.xls")
))

read_1970s_sheet <- function(year) {
  raw <- suppressMessages(read_excel(
    file.path(data_raw_dir, "pe-11-1970s.xls"),
    sheet = as.character(year),
    col_names = FALSE,
    col_types = "text"
  ))
  tibble(
    Year = year,
    AgeRaw = raw[[1]],
    `All races.Total` = raw[[2]],
    `All races.Male` = raw[[3]],
    `All races.Female` = raw[[4]],
    `White.Total` = raw[[5]],
    `White.Male` = raw[[6]],
    `White.Female` = raw[[7]],
    `Black.Total` = raw[[8]],
    `Black.Male` = raw[[9]],
    `Black.Female` = raw[[10]]
  ) |>
    filter(grepl("^[0-9]+$", AgeRaw) | AgeRaw == "85+") |>
    mutate(
      Age = as.numeric(if_else(AgeRaw == "85+", "85", AgeRaw)),
      across(`All races.Total`:`Black.Female`, as.numeric)
    ) |>
    select(Year, Age, `All races.Total`:`Black.Female`)
}

Census_1970s_by_age <- bind_rows(lapply(1970:1979, read_1970s_sheet))

cat("Years extracted:", paste(sort(unique(Census_1970s_by_age$Year)), collapse = ", "))
```

```{r aggregate-1970s, dependson=c("extract-1970s-from-xls", "aggregate-1960s")}
Census_1970s_totals_adults <- aggregate_3cat(Census_1970s_by_age |> filter(Age >= 18))
Census_1970s_totals_allages <- aggregate_3cat(Census_1970s_by_age)

Census_1970s_adults <- add_race3_and_race2(reshape_3cat(Census_1970s_totals_adults))
Census_1970s_allages <- add_race3_and_race2(reshape_3cat(Census_1970s_totals_allages))
```

The check below verifies that Male and Female sum to the reported total and
that White, Black, and Other sum to the reported total, both of which
should be exactly zero.

```{r check-1970s, dependson="aggregate-1970s"}
Census_1970s_totals_adults |>
  mutate(
    Diff.MaleFemale = `All races.Total` - `All races.Male` - `All races.Female`,
    Diff.RaceSum = `All races.Total` - `White.Total` - `Black.Total` - `Other.Total`
  ) |>
  summarize(MaxAbsDiff.MaleFemale = max(abs(Diff.MaleFemale)), MaxAbsDiff.RaceSum = max(abs(Diff.RaceSum)))
```

### 1980-1989: 1980s national intercensal RQI files

10 quarterly national files, one per year-pair, from the Census Bureau's
["National Intercensal Datasets: 1980-1990"](https://www.census.gov/data/datasets/time-series/demo/popest/1980s-national.html)
(distributed as zips, e.g. `e8081rqi.zip` containing `E8081RQI.TXT`), laid
out per the
[file-layout documentation](https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/1980-1990/nat-detail-layout.txt).
First source with a Hispanic-origin breakdown, fully crossed with race and
sex, and the first source with `race4`-level race detail (White, Black,
American Indian/Alaska Native, and Asian/Pacific Islander reported
separately). This section expects all 10 years from 1980 through 1989. The
years listed below confirm every year was found with no gaps.

```{r load-RQI-files, dependson=c("setup", "download-if-missing")}
rqi_dir <- file.path(data_raw_dir, "1980-1989")
dir.create(rqi_dir, showWarnings = FALSE, recursive = TRUE)

download_RQI_txt <- function(year) {
  yy <- year %% 100
  yy2 <- (year + 1) %% 100
  code <- sprintf("%02d%02d", yy, yy2)
  txt_path <- file.path(rqi_dir, sprintf("E%sRQI.TXT", code))
  if (!file.exists(txt_path)) {
    zip_path <- file.path(rqi_dir, sprintf("e%srqi.zip", code))
    download_if_missing(
      sprintf("https://www2.census.gov/programs-surveys/popest/datasets/1980-1990/national/asrh/e%srqi.zip", code),
      zip_path
    )
    unzip(zip_path, exdir = rqi_dir)
  }
  txt_path
}

rqi_col_names <- c(
  "Series", "Month", "Year", "Age", "Blank",
  "Total", "Total.Male", "Total.Female",
  "White.Male", "White.Female",
  "Black.Male", "Black.Female",
  "AIAE.Male", "AIAE.Female",
  "API.Male", "API.Female",
  "Hispanic.Male", "Hispanic.Female",
  "White.NonHispanic.Male", "White.NonHispanic.Female",
  "Black.NonHispanic.Male", "Black.NonHispanic.Female",
  "AIAE.NonHispanic.Male", "AIAE.NonHispanic.Female",
  "API.NonHispanic.Male", "API.NonHispanic.Female"
)

load_RQI_year <- function(year) {
  yy <- year %% 100
  path <- download_RQI_txt(year)
  read_fwf(
    path,
    fwf_widths(c(2, 2, 2, 4, 1, rep(10, 21)), col_names = rqi_col_names),
    show_col_types = FALSE
  ) |>
    filter(Month == 7, Year == yy, Age <= 100) |>
    mutate(SourceYear = year)
}

years_1980s <- 1980:1989

RQI_raw <- bind_rows(lapply(years_1980s, load_RQI_year))

cat("Years extracted:", paste(sort(unique(RQI_raw$SourceYear)), collapse = ", "))
```

```{r aggregate-RQI, dependson="load-RQI-files"}
aggregate_RQI <- function(raw_data) {
  raw_data |>
    group_by(SourceYear) |>
    summarize(across(Total:API.NonHispanic.Female, sum), .groups = "drop") |>
    rename(Year = SourceYear)
}

RQI_agg_adults <- aggregate_RQI(RQI_raw |> filter(Age >= 18))
RQI_agg_allages <- aggregate_RQI(RQI_raw)
```

The check below verifies that the four race categories sum to the reported
total and that the Hispanic and non-Hispanic race categories also sum to
the reported total, both of which should be exactly zero.

```{r check-RQI, dependson="aggregate-RQI"}
RQI_agg_adults |>
  mutate(
    diff_race = Total -
      (White.Male + White.Female + Black.Male + Black.Female +
         AIAE.Male + AIAE.Female + API.Male + API.Female),
    diff_hisp = Total -
      (Hispanic.Male + Hispanic.Female +
         White.NonHispanic.Male + White.NonHispanic.Female +
         Black.NonHispanic.Male + Black.NonHispanic.Female +
         AIAE.NonHispanic.Male + AIAE.NonHispanic.Female +
         API.NonHispanic.Male + API.NonHispanic.Female)
  ) |>
  summarize(MaxAbsDiff.Race = max(abs(diff_race)), MaxAbsDiff.Hispanic = max(abs(diff_hisp)))
```

```{r reshape-RQI, dependson=c("aggregate-RQI", "race-scheme-collapse-maps")}
# Race labels renamed AIAE -> AIAN here to match this file's race6 naming
# for the same group in later eras (see race-scheme-collapse-maps above).
reshape_RQI4 <- function(agg) {
  agg |>
    transmute(
      Year,
      White.Male, White.Female,
      Black.Male, Black.Female,
      AIAN.Male = AIAE.Male, AIAN.Female = AIAE.Female,
      API.Male, API.Female
    ) |>
    pivot_longer(
      cols = -Year,
      names_to = c("Race", "Sex"),
      names_sep = "\\.",
      values_to = "Population"
    )
}

# Tags a race4 (White/Black/AIAN/API) table with RaceScheme and adds its
# race3- and race2-collapsed derivatives, for eras whose native scheme is
# race4. Works for both the plain Race x Sex tables and the Race x
# HispanicOrigin x Sex crosses, since collapse_race() groups by every
# column except Population.
add_race4_race3_race2 <- function(race4_table) {
  race4 <- race4_table |> mutate(RaceScheme = "race4")
  bind_rows(
    race4,
    collapse_race(race4, race4_to_race3, "race3"),
    collapse_race(race4, race4_to_race2, "race2")
  )
}

# Same as add_race4_race3_race2(), for the "18-19 exact split" tables,
# whose value column is Population_1819 instead of Population.
add_race4_race3_race2_1819 <- function(race4_table) {
  race4 <- race4_table |> mutate(RaceScheme = "race4")
  bind_rows(
    race4,
    collapse_race_1819(race4, race4_to_race3, "race3"),
    collapse_race_1819(race4, race4_to_race2, "race2")
  )
}

Census_1980s_adults <- add_race4_race3_race2(reshape_RQI4(RQI_agg_adults))
Census_1980s_allages <- add_race4_race3_race2(reshape_RQI4(RQI_agg_allages))

# Full Race x HispanicOrigin x Sex cross, ages 18+, kept at race4
# granularity: the RQI file's AIAE.NonHispanic/API.NonHispanic fields are
# separate columns, so a Hispanic split is available for AIAN and API
# individually, not just for the collapsed race3/race2 view.
reshape_RQI4_hispanic <- function(agg) {
  agg |>
    transmute(
      Year,
      White.Hispanic.Male = White.Male - White.NonHispanic.Male,
      White.Hispanic.Female = White.Female - White.NonHispanic.Female,
      White.NonHispanic.Male, White.NonHispanic.Female,
      Black.Hispanic.Male = Black.Male - Black.NonHispanic.Male,
      Black.Hispanic.Female = Black.Female - Black.NonHispanic.Female,
      Black.NonHispanic.Male, Black.NonHispanic.Female,
      AIAN.Hispanic.Male = AIAE.Male - AIAE.NonHispanic.Male,
      AIAN.Hispanic.Female = AIAE.Female - AIAE.NonHispanic.Female,
      AIAN.NonHispanic.Male = AIAE.NonHispanic.Male,
      AIAN.NonHispanic.Female = AIAE.NonHispanic.Female,
      API.Hispanic.Male = API.Male - API.NonHispanic.Male,
      API.Hispanic.Female = API.Female - API.NonHispanic.Female,
      API.NonHispanic.Male, API.NonHispanic.Female
    ) |>
    pivot_longer(
      cols = -Year,
      names_to = c("Race", "HispanicOrigin", "Sex"),
      names_sep = "\\.",
      values_to = "Population"
    )
}

Census_1980s_hispanic_adults <- add_race4_race3_race2(reshape_RQI4_hispanic(RQI_agg_adults))
```

### 1990-1999: 1990-2000 Intercensal (county-summed)

10 direct flat-file downloads, `stch-icen1990.txt` through `stch-icen1999.txt`
(one row per County x AgeGroup x RaceSex x Ethnicity), from the Census
Bureau's
[1990-2000 intercensal county file directory](https://www2.census.gov/programs-surveys/popest/tables/1990-2000/intercensal/st-co/),
laid out per the
[file-layout documentation](https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/1990-2000/stch-intercensal_layout.txt).
This product's `RACE_SEX` variable is `race4`-level (White, Black, American
Indian/Alaska Native, and Asian/Pacific Islander, each crossed with sex),
and its `AGEGRP`/`HISP` scheme has no single-year-of-age detail or
pre-built 18+ aggregate, so the 18+ total is approximated as 2/5 of the "15
to 19 years" bin (assuming a uniform within-bin age distribution) plus the
20+ bins in full -- the one approximate figure in this table. This section
expects all 10 years from 1990 through 1999. The years listed below confirm
every year was found with no gaps.

```{r load-1990s-county-data, dependson=c("setup", "download-if-missing")}
census_1990s_flat_dir <- file.path(data_raw_dir, "1990-1999-county")
dir.create(census_1990s_flat_dir, showWarnings = FALSE, recursive = TRUE)

for (year in 1990:1999) {
  download_if_missing(
    sprintf("https://www2.census.gov/programs-surveys/popest/tables/1990-2000/intercensal/st-co/stch-icen%d.txt", year),
    file.path(census_1990s_flat_dir, sprintf("stch-icen%d.txt", year))
  )
}

census_1990s_col_widths <- fwf_widths(
  c(2, 2, 5, 3, 2, 2, 7),
  col_names = c("YEAR", "Blank", "StateCounty", "AGEGRP", "RACE_SEX", "HISP", "POP")
)

load_1990s_flat_year <- function(year) {
  path <- file.path(census_1990s_flat_dir, sprintf("stch-icen%d.txt", year))
  read_fwf(path, census_1990s_col_widths, col_types = "iiciiii", show_col_types = FALSE) |>
    mutate(YEAR = year - 1900) |>
    group_by(YEAR, AGEGRP, RACE_SEX, HISP) |>
    summarize(POP = sum(POP), .groups = "drop")
}

census_1990s_national <- bind_rows(lapply(1990:1999, load_1990s_flat_year)) |>
  mutate(Year = 1900 + YEAR)

cat("Years extracted:", paste(sort(unique(census_1990s_national$Year)), collapse = ", "))
```

```{r derive-1990s-race-sex, dependson=c("load-1990s-county-data", "race-scheme-collapse-maps", "reshape-RQI")}
census_1990s_race_sex_hisp_raw <- census_1990s_national |>
  mutate(
    Sex = if_else(RACE_SEX %% 2 == 1, "Male", "Female"),
    Race = case_when(
      RACE_SEX %in% c(1, 2) ~ "White",
      RACE_SEX %in% c(3, 4) ~ "Black",
      RACE_SEX %in% c(5, 6) ~ "AIAN",
      RACE_SEX %in% c(7, 8) ~ "API"
    ),
    HispanicOrigin = if_else(HISP == 2, "Hispanic", "NonHispanic")
  )

census_1990s_race4_allages <- census_1990s_race_sex_hisp_raw |>
  group_by(Year, Sex, Race) |>
  summarize(Population = sum(POP), .groups = "drop")
census_1990s_allages <- add_race4_race3_race2(census_1990s_race4_allages)

# 18+ approximation: AGEGRP 5-18 (ages 20+) in full, plus 2/5 of AGEGRP 4
# ("15 to 19 years"), assuming a uniform within-bin age distribution.
census_1990s_race4_adults <- census_1990s_race_sex_hisp_raw |>
  filter(AGEGRP >= 4) |>
  mutate(Weight = if_else(AGEGRP == 4, 0.4, 1)) |>
  group_by(Year, Sex, Race) |>
  summarize(Population = sum(POP * Weight), .groups = "drop")
census_1990s_adults <- add_race4_race3_race2(census_1990s_race4_adults)

census_1990s_race4_hispanic_adults <- census_1990s_race_sex_hisp_raw |>
  filter(AGEGRP >= 4) |>
  mutate(Weight = if_else(AGEGRP == 4, 0.4, 1)) |>
  group_by(Year, Sex, Race, HispanicOrigin) |>
  summarize(Population = sum(POP * Weight), .groups = "drop")
census_1990s_hispanic_adults <- add_race4_race3_race2(census_1990s_race4_hispanic_adults)
```

The check below verifies that, for every RaceScheme, the Race x Sex
marginal equals the Race x Sex x HispanicOrigin cross summed over
HispanicOrigin. This should be exactly zero.

```{r check-1990s, dependson="derive-1990s-race-sex"}
census_1990s_hispanic_adults |>
  group_by(Year, RaceScheme, Sex, Race) |>
  summarize(PopulationFromCross = sum(Population), .groups = "drop") |>
  inner_join(census_1990s_adults, by = c("Year", "RaceScheme", "Sex", "Race")) |>
  mutate(Diff = PopulationFromCross - Population) |>
  summarize(MaxAbsDiff = max(abs(Diff)))
```

### 2000-2019: two intercensal products

2000-2010 Intercensal (`censusapi`, `pep/int_charagegroups`, vintage 2000)
for 2000-2009, and 2010-2020 Intercensal (`NC-EST2020INT-ASR6H`, a
downloaded workbook) for 2010-2019 -- both benchmarked against two
decennial censuses, unlike the postcensal alternatives this pass avoids
(postcensal products are built forward from a single census with no later
correction; see the source EDA's "Why intercensal, not postcensal" sections
for the two cases where that distinction mattered). Both sources report
`race6`-level race detail (White, Black, American Indian/Alaska Native,
Asian, Native Hawaiian/Pacific Islander, and Two or More Races). This
section expects all 10 years from 2000 through 2009. The years listed below
confirm every year was found with no gaps.

```{r load-2000s-national-data, dependson="setup"}
census_2000s_raw <- getCensus(
  name = "pep/int_charagegroups",
  vintage = 2000,
  vars = c("POP", "DATE_DESC", "AGEGROUP", "RACE", "SEX", "HISP"),
  region = "us:1"
) |>
  mutate(across(c(POP, AGEGROUP, RACE, SEX, HISP), as.integer)) |>
  filter(grepl("^7/1/[0-9]{4} population estimate$", DATE_DESC)) |>
  mutate(Year = as.integer(sub("^7/1/([0-9]{4}).*$", "\\1", DATE_DESC)))

cat("2000s years extracted:", paste(sort(unique(census_2000s_raw$Year)), collapse = ", "))
```

This section expects the 10 years from 2010 through 2019. The years listed
below confirm every year was found with no gaps.

```{r load-2010s-national-data, dependson=c("setup", "download-if-missing")}
invisible(download_if_missing(
  "https://www2.census.gov/programs-surveys/popest/tables/2010-2020/intercensal/national/asrh/nc-est2020int-asr6h.xlsx",
  file.path(data_raw_dir, "nc-est2020int-asr6h.xlsx")
))

# Row map: each row gives one (Sex, HispanicOrigin, AgeGroup) combination,
# coded to match this table's SEX (1/2 = Male/Female), HISP (0/1/2 =
# Both/NotHispanic/Hispanic), and AGEGROUP (0/29 = all ages/18+) scheme.
intercensal_2010s_row_map <- tribble(
  ~Sex, ~Hisp, ~AgeGroup, ~Row, ~Label,
  1, 0,  0,  42, ".MALE",
  1, 0, 29,  74, "..18 years and over",
  2, 0,  0,  78, ".FEMALE",
  2, 0, 29, 110, "..18 years and over",
  1, 1,  0, 150, "..MALE",
  1, 1, 29, 182, "...18 years and over",
  2, 1,  0, 186, "..FEMALE",
  2, 1, 29, 218, "...18 years and over",
  1, 2,  0, 258, "..MALE",
  1, 2, 29, 290, "...18 years and over",
  2, 2,  0, 294, "..FEMALE",
  2, 2, 29, 326, "...18 years and over"
)

# Row map for the 18 five-year AgeGroup bins (AGEGROUP 1-18), matching this
# table's 2000-2009 AGEGROUP coding exactly.
age_group_source_labels_2010s <- c(
  "Under 5 years", "5 to 9 years", "10 to 14 years", "15 to 19 years",
  "20 to 24 years", "25 to 29 years", "30 to 34 years", "35 to 39 years",
  "40 to 44 years", "45 to 49 years", "50 to 54 years", "55 to 59 years",
  "60 to 64 years", "65 to 69 years", "70 to 74 years", "75 to 79 years",
  "80 to 84 years", "85 years and over"
)

intercensal_2010s_age_blocks <- tribble(
  ~Sex, ~Hisp, ~StartRow, ~Dots,
  1, 0,  43, 2,  # MALE
  2, 0,  79, 2,  # FEMALE
  1, 1, 151, 3,  # NOT HISPANIC..MALE
  2, 1, 187, 3,  # NOT HISPANIC..FEMALE
  1, 2, 259, 3,  # HISPANIC..MALE
  2, 2, 295, 3   # HISPANIC..FEMALE
)

intercensal_2010s_age_row_map <- bind_rows(lapply(seq_len(nrow(intercensal_2010s_age_blocks)), function(i) {
  b <- intercensal_2010s_age_blocks[i, ]
  tibble(
    Sex = b$Sex,
    Hisp = b$Hisp,
    AgeGroup = 1:18,
    Row = b$StartRow + 0:17,
    Label = paste0(strrep(".", b$Dots), age_group_source_labels_2010s)
  )
}))

intercensal_2010s_row_map <- bind_rows(intercensal_2010s_row_map, intercensal_2010s_age_row_map)

intercensal_2010s_race_codes <- tribble(
  ~Sheet,                ~RACE,
  "White",                1L,
  "Black",                2L,
  "AIAN",                 3L,
  "Asian",                4L,
  "NHPI",                 5L,
  "Two or More Races",    6L
)

intercensal_2010s_year_cols <- 3:12
intercensal_2010s_years <- 2010:2019

read_intercensal_2010s_sheet <- function(sheet_name, race_code) {
  raw <- suppressMessages(read_excel(
    file.path(data_raw_dir, "nc-est2020int-asr6h.xlsx"),
    sheet = sheet_name,
    col_names = FALSE,
    col_types = "text"
  ))

  # Fail loudly, not silently, if a future re-download of this file ever
  # shifts the row layout assumed above.
  mismatches <- intercensal_2010s_row_map |>
    rowwise() |>
    mutate(Actual = trimws(raw[[1]][Row])) |>
    ungroup() |>
    filter(Actual != Label)
  if (nrow(mismatches) > 0) {
    stop(sprintf(
      "Row label mismatch in sheet '%s': %s",
      sheet_name,
      paste(sprintf("row %d expected '%s' got '%s'", mismatches$Row, mismatches$Label, mismatches$Actual), collapse = "; ")
    ))
  }

  bind_rows(lapply(seq_len(nrow(intercensal_2010s_row_map)), function(i) {
    tibble(
      Year = intercensal_2010s_years,
      POP = as.numeric(unlist(raw[intercensal_2010s_row_map$Row[i], intercensal_2010s_year_cols], use.names = FALSE)),
      SEX = intercensal_2010s_row_map$Sex[i],
      HISP = intercensal_2010s_row_map$Hisp[i],
      AGEGROUP = intercensal_2010s_row_map$AgeGroup[i]
    )
  })) |>
    mutate(RACE = race_code)
}

census_2010s_raw <- bind_rows(Map(
  read_intercensal_2010s_sheet,
  intercensal_2010s_race_codes$Sheet,
  intercensal_2010s_race_codes$RACE
))

cat("2010s years extracted:", paste(sort(unique(census_2010s_raw$Year)), collapse = ", "))
```

```{r combine-2000s-2010s, dependson=c("load-2000s-national-data", "load-2010s-national-data", "race-scheme-collapse-maps")}
census_2000_2019_raw <- bind_rows(
  census_2000s_raw |> select(Year, POP, AGEGROUP, RACE, SEX, HISP),
  census_2010s_raw |> select(Year, POP, AGEGROUP, RACE, SEX, HISP)
)

race_labels_2000s <- c(
  `1` = "White", `2` = "Black", `3` = "AIAN",
  `4` = "Asian", `5` = "NHPI", `6` = "TwoOrMore"
)

census_2000_2019_race_sex_raw <- census_2000_2019_raw |>
  filter(RACE %in% 1:6, SEX %in% 1:2) |>
  mutate(
    Race = race_labels_2000s[as.character(RACE)],
    Sex = if_else(SEX == 1, "Male", "Female")
  )

# Tags a race6 (White/Black/AIAN/Asian/NHPI/TwoOrMore) table with
# RaceScheme and adds its race3- and race2-collapsed derivatives. No race4
# derivative: Two or More Races has no home in the four-category scheme
# (see "Overview" above), so race4 stays native-only, 1980-1999.
add_race6_race3_race2 <- function(race6_table) {
  race6 <- race6_table |> mutate(RaceScheme = "race6")
  bind_rows(
    race6,
    collapse_race(race6, race6_to_race3, "race3"),
    collapse_race(race6, race6_to_race2, "race2")
  )
}

# Same as add_race6_race3_race2(), for the "18-19 exact/interpolated
# split" tables, whose value column is Population_1819 instead of
# Population.
add_race6_race3_race2_1819 <- function(race6_table) {
  race6 <- race6_table |> mutate(RaceScheme = "race6")
  bind_rows(
    race6,
    collapse_race_1819(race6, race6_to_race3, "race3"),
    collapse_race_1819(race6, race6_to_race2, "race2")
  )
}

census_2000_2019_race6_allages <- census_2000_2019_race_sex_raw |>
  filter(AGEGROUP == 0, HISP == 0) |>
  group_by(Year, Sex, Race) |>
  summarize(Population = sum(POP), .groups = "drop")
census_2000_2019_allages <- add_race6_race3_race2(census_2000_2019_race6_allages)

census_2000_2019_race6_adults <- census_2000_2019_race_sex_raw |>
  filter(AGEGROUP == 29, HISP == 0) |>
  group_by(Year, Sex, Race) |>
  summarize(Population = sum(POP), .groups = "drop")
census_2000_2019_adults <- add_race6_race3_race2(census_2000_2019_race6_adults)

census_2000_2019_race6_hispanic_adults <- census_2000_2019_race_sex_raw |>
  filter(AGEGROUP == 29, HISP %in% 1:2) |>
  mutate(HispanicOrigin = if_else(HISP == 2, "Hispanic", "NonHispanic")) |>
  group_by(Year, Sex, Race, HispanicOrigin) |>
  summarize(Population = sum(POP), .groups = "drop")
census_2000_2019_hispanic_adults <- add_race6_race3_race2(census_2000_2019_race6_hispanic_adults)
```

The check below verifies that, for every RaceScheme, the "Both" Hispanic
total (`HISP == 0`) equals the sum of the Hispanic and non-Hispanic totals.
This should be exactly zero.

```{r check-2000-2019, dependson="combine-2000s-2010s"}
census_2000_2019_hispanic_adults |>
  group_by(Year, RaceScheme, Sex, Race) |>
  summarize(PopulationFromCross = sum(Population), .groups = "drop") |>
  inner_join(census_2000_2019_adults, by = c("Year", "RaceScheme", "Sex", "Race")) |>
  mutate(Diff = PopulationFromCross - Population) |>
  summarize(MaxAbsDiff = max(abs(Diff)))
```

### 2020-2025: post-2020 vintage

14 monthly national files (`nc-est2025-alldata-c-file{01..14}.csv`,
covering April 2020 through July 2025 plus short-term projections to
December 2026), from the Census Bureau's
["National Population by Characteristics: 2020-2025"](https://www.census.gov/data/datasets/time-series/demo/popest/2020s-national-detail.html)
datasets page (Vintage 2025 files, directly downloadable at
`https://www2.census.gov/programs-surveys/popest/datasets/2020-2025/national/asrh/`).
This file reports `race6`-level race detail directly (`WA`/`BA`/`IA`/`AA`/
`NA`/`TOM` for White/Black/AIAN/Asian/NHPI/Two-or-More Alone), each also
crossed with Hispanic origin (`H`/`NH` prefixes), so no category needs to
be computed as a residual here. Unlike Vintage 2024's 5 files (one per July
2020-2024), Vintage 2025's file numbering doesn't map 1:1 to specific
months, so all 14 files are downloaded and combined, then filtered down to
July estimates for 2020-2025. Note that each new PEP vintage revises the
entire postcensal time series back to the last census, so this also
replaces (not supplements) the previously-published Vintage 2024 figures
for 2020-2024. This section expects the 6 years from 2020 through 2025.
The years listed below confirm every year was found with no gaps.

```{r load-2020s-census-files, dependson=c("setup", "download-if-missing")}
census_2020s_dir <- file.path(data_raw_dir, "2020-2025")
dir.create(census_2020s_dir, showWarnings = FALSE, recursive = TRUE)

census_2020s_files <- tibble(File = sprintf("file%02d", 1:14))

for (f in census_2020s_files$File) {
  download_if_missing(
    sprintf("https://www2.census.gov/programs-surveys/popest/datasets/2020-2025/national/asrh/nc-est2025-alldata-c-%s.csv", f),
    file.path(census_2020s_dir, sprintf("nc-est2025-alldata-c-%s.csv", f))
  )
}

census_2020s_raw <- census_2020s_files$File |>
  lapply(\(f) {
    read_csv(
      file.path(census_2020s_dir, paste0('nc-est2025-alldata-c-', f, '.csv')),
      show_col_types = FALSE
    )
  }) |>
  bind_rows() |>
  filter(MONTH == 7, YEAR %in% 2020:2025)

cat("Years extracted:", paste(sort(unique(census_2020s_raw$YEAR)), collapse = ", "))
```

```{r summarize-2020s-race-sex, dependson=c('load-2020s-census-files', 'combine-2000s-2010s')}
summarize_2020s_race_sex <- function(data) {
  data |>
    group_by(YEAR) |>
    summarize(
      White_Male = sum(WA_MALE), White_Female = sum(WA_FEMALE),
      Black_Male = sum(BA_MALE), Black_Female = sum(BA_FEMALE),
      AIAN_Male = sum(IA_MALE), AIAN_Female = sum(IA_FEMALE),
      Asian_Male = sum(AA_MALE), Asian_Female = sum(AA_FEMALE),
      NHPI_Male = sum(NA_MALE), NHPI_Female = sum(NA_FEMALE),
      TwoOrMore_Male = sum(TOM_MALE), TwoOrMore_Female = sum(TOM_FEMALE),
      .groups = 'drop'
    ) |>
    pivot_longer(
      cols = -YEAR,
      names_to = c('Race', 'Sex'),
      names_sep = '_',
      values_to = 'Population'
    ) |>
    rename(Year = YEAR) |>
    arrange(Year, Race, Sex)
}

census_2020s_race6_allages <- census_2020s_raw |>
  filter(AGE == 999) |>
  summarize_2020s_race_sex()
census_2020s_race_sex_allages <- add_race6_race3_race2(census_2020s_race6_allages)

census_2020s_race6_adults <- census_2020s_raw |>
  filter(AGE != 999, AGE >= 18) |>
  summarize_2020s_race_sex()
census_2020s_race_sex_adults <- add_race6_race3_race2(census_2020s_race6_adults)
```

```{r summarize-2020s-hispanic, dependson=c('load-2020s-census-files', 'combine-2000s-2010s')}
summarize_2020s_race_hispanic <- function(data) {
  data |>
    group_by(YEAR) |>
    summarize(
      White.Hispanic_Male = sum(HWA_MALE), White.Hispanic_Female = sum(HWA_FEMALE),
      White.NonHispanic_Male = sum(NHWA_MALE), White.NonHispanic_Female = sum(NHWA_FEMALE),
      Black.Hispanic_Male = sum(HBA_MALE), Black.Hispanic_Female = sum(HBA_FEMALE),
      Black.NonHispanic_Male = sum(NHBA_MALE), Black.NonHispanic_Female = sum(NHBA_FEMALE),
      AIAN.Hispanic_Male = sum(HIA_MALE), AIAN.Hispanic_Female = sum(HIA_FEMALE),
      AIAN.NonHispanic_Male = sum(NHIA_MALE), AIAN.NonHispanic_Female = sum(NHIA_FEMALE),
      Asian.Hispanic_Male = sum(HAA_MALE), Asian.Hispanic_Female = sum(HAA_FEMALE),
      Asian.NonHispanic_Male = sum(NHAA_MALE), Asian.NonHispanic_Female = sum(NHAA_FEMALE),
      NHPI.Hispanic_Male = sum(HNA_MALE), NHPI.Hispanic_Female = sum(HNA_FEMALE),
      NHPI.NonHispanic_Male = sum(NHNA_MALE), NHPI.NonHispanic_Female = sum(NHNA_FEMALE),
      TwoOrMore.Hispanic_Male = sum(HTOM_MALE), TwoOrMore.Hispanic_Female = sum(HTOM_FEMALE),
      TwoOrMore.NonHispanic_Male = sum(NHTOM_MALE), TwoOrMore.NonHispanic_Female = sum(NHTOM_FEMALE),
      .groups = 'drop'
    ) |>
    pivot_longer(
      cols = -YEAR,
      names_to = c('Race', 'HispanicOrigin', 'Sex'),
      names_sep = '[._]',
      values_to = 'Population'
    ) |>
    rename(Year = YEAR) |>
    arrange(Year, Race, HispanicOrigin, Sex)
}

census_2020s_race6_hispanic_adults <- census_2020s_raw |>
  filter(AGE != 999, AGE >= 18) |>
  summarize_2020s_race_hispanic()
census_2020s_hispanic_adults <- add_race6_race3_race2(census_2020s_race6_hispanic_adults)
```

The check below verifies that, for every RaceScheme, the Race x Sex
marginal equals the Race x Sex x HispanicOrigin cross summed over
HispanicOrigin. This should be exactly zero.

```{r check-2020s, dependson=c("summarize-2020s-race-sex", "summarize-2020s-hispanic")}
census_2020s_hispanic_adults |>
  group_by(Year, RaceScheme, Sex, Race) |>
  summarize(PopulationFromCross = sum(Population), .groups = "drop") |>
  inner_join(census_2020s_race_sex_adults, by = c("Year", "RaceScheme", "Sex", "Race")) |>
  mutate(Diff = PopulationFromCross - Population) |>
  summarize(MaxAbsDiff = max(abs(Diff)))
```

## Combined Target Table

Stacks every era's Year x RaceScheme x Race x Sex table into one 1960-2025
range (1900-1959 kept separate -- see "Overview" above), plus the
1980-2025 Hispanic-origin cross. Every era's tables already carry a
RaceScheme column from the sections above, so this stacking step needs no
further collapsing.

```{r combine-all-eras, dependson=c("aggregate-1960s", "aggregate-1970s", "reshape-RQI", "derive-1990s-race-sex", "combine-2000s-2010s", "summarize-2020s-race-sex")}
census_targets_allages <- bind_rows(
  Census_1960s_allages, Census_1970s_allages, Census_1980s_allages,
  census_1990s_allages, census_2000_2019_allages, census_2020s_race_sex_allages
)

census_targets_adults <- bind_rows(
  Census_1960s_adults, Census_1970s_adults, Census_1980s_adults,
  census_1990s_adults, census_2000_2019_adults, census_2020s_race_sex_adults
)
```

```{r combine-1900-1959, dependson="aggregate-1900-1959"}
census_targets_1900_1959_allages <- Census_1900_1959_allages
census_targets_1900_1959_adults <- Census_1900_1959_adults
```

```{r combine-hispanic, dependson=c("reshape-RQI", "derive-1990s-race-sex", "combine-2000s-2010s", "summarize-2020s-hispanic")}
census_targets_hispanic_adults <- bind_rows(
  Census_1980s_hispanic_adults, census_1990s_hispanic_adults,
  census_2000_2019_hispanic_adults, census_2020s_hispanic_adults
)
```

The check below verifies that, for every RaceScheme, Hispanic plus
NonHispanic sums to the Race x Sex total built independently above, for
every 1980-2025 year. This should be exactly zero.

```{r check-combine-hispanic, dependson="combine-hispanic"}
census_targets_hispanic_adults |>
  group_by(Year, RaceScheme, Race, Sex) |>
  summarize(PopulationFromHispanicCross = sum(Population), .groups = "drop") |>
  inner_join(census_targets_adults, by = c("Year", "RaceScheme", "Race", "Sex")) |>
  mutate(Diff = PopulationFromHispanicCross - Population) |>
  summarize(MaxAbsDiff = max(abs(Diff)))
```

### Age Group Detail

Bins each era's already-extracted single-year-of-age (or, for 1980-2025,
already-crossed) data into the 18 `age_group_levels` bins, then checks that
each era's by-age table reconciles exactly with its own already-verified
all-ages and adults totals above, for every RaceScheme that era supports.
The 18+ cutoff splits the "15 to 19" bin, so the adults check needs an
exact (or, for 1990-1999 and 2000-2019, the documented 2/5-interpolated)
18-19 count per era and RaceScheme.

```{r age-group-1900-1959, dependson=c("extract-1900-1959-from-xls", "aggregate-1900-1959", "bin-age-5yr", "check-allages-reconciliation", "check-adults-reconciliation")}
census_targets_1900_1959_by_age <- bind_rows(
  Census_1900_1959_by_age |> filter(Year < 1940) |> mutate(AgeGroup = bin_age_5yr(Age, top_bin_age = 75)),
  Census_1900_1959_by_age |> filter(Year >= 1940) |> mutate(AgeGroup = bin_age_5yr(Age, top_bin_age = 85))
) |>
  pivot_longer(
    cols = `White.Total`:`Nonwhite.Female`,
    names_to = c("Race", "Sex"),
    names_sep = "\\.",
    values_to = "Population"
  ) |>
  filter(Sex %in% c("Male", "Female")) |>
  group_by(Year, Race, Sex, AgeGroup) |>
  summarize(Population = sum(Population), .groups = "drop") |>
  mutate(RaceScheme = "race2")

age_1900_1959_18_19 <- reshape_2cat(aggregate_2cat(Census_1900_1959_by_age |> filter(Age %in% c(18, 19)))) |>
  rename(Population_1819 = Population) |>
  mutate(RaceScheme = "race2")
```

The check below verifies that this era's by-age table reconciles with its
all-ages and adults totals. Both rows should be exactly zero.

```{r check-age-group-1900-1959, dependson="age-group-1900-1959"}
bind_rows(
  AllAges = check_allages_reconciliation(
    census_targets_1900_1959_by_age, census_targets_1900_1959_allages, c("Year", "RaceScheme", "Race", "Sex")
  ),
  Adults = check_adults_reconciliation(
    census_targets_1900_1959_by_age, age_1900_1959_18_19, census_targets_1900_1959_adults, c("Year", "RaceScheme", "Race", "Sex")
  ),
  .id = "Check"
)
```

```{r age-group-1960s, dependson=c("extract-1960s-from-xls", "aggregate-1960s", "bin-age-5yr", "check-allages-reconciliation", "check-adults-reconciliation")}
age_1960s_all <- Census_1960s_by_age |>
  mutate(AgeGroup = bin_age_5yr(Age)) |>
  group_by(Year, AgeGroup) |>
  summarize(across(`All races.Total`:`Black.Female`, sum), .groups = "drop") |>
  mutate(
    Other.Male = `All races.Male` - White.Male - Black.Male,
    Other.Female = `All races.Female` - White.Female - Black.Female
  )

census_targets_1960s_by_age <- age_1960s_all |>
  select(Year, AgeGroup, starts_with("White."), starts_with("Black."), starts_with("Other.")) |>
  pivot_longer(
    cols = -c(Year, AgeGroup),
    names_to = c("Race", "Sex"),
    names_sep = "\\.",
    values_to = "Population"
  ) |>
  filter(Sex %in% c("Male", "Female")) |>
  add_race3_and_race2()

age_1960s_18_19 <- reshape_3cat(aggregate_3cat(Census_1960s_by_age |> filter(Age %in% c(18, 19)))) |>
  rename(Population_1819 = Population) |>
  add_race3_and_race2_1819()
```

The check below verifies that this era's by-age table reconciles with its
all-ages and adults totals, for both RaceScheme values. All four rows
should be exactly zero.

```{r check-age-group-1960s, dependson="age-group-1960s"}
bind_rows(
  AllAges = check_allages_reconciliation(
    census_targets_1960s_by_age, Census_1960s_allages, c("Year", "RaceScheme", "Race", "Sex")
  ),
  Adults = check_adults_reconciliation(
    census_targets_1960s_by_age, age_1960s_18_19, Census_1960s_adults, c("Year", "RaceScheme", "Race", "Sex")
  ),
  .id = "Check"
)
```

```{r age-group-1970s, dependson=c("extract-1970s-from-xls", "aggregate-1970s", "aggregate-1960s", "bin-age-5yr", "check-allages-reconciliation", "check-adults-reconciliation")}
age_1970s_all <- Census_1970s_by_age |>
  mutate(AgeGroup = bin_age_5yr(Age)) |>
  group_by(Year, AgeGroup) |>
  summarize(across(`All races.Total`:`Black.Female`, sum), .groups = "drop") |>
  mutate(
    Other.Male = `All races.Male` - White.Male - Black.Male,
    Other.Female = `All races.Female` - White.Female - Black.Female
  )

census_targets_1970s_by_age <- age_1970s_all |>
  select(Year, AgeGroup, starts_with("White."), starts_with("Black."), starts_with("Other.")) |>
  pivot_longer(
    cols = -c(Year, AgeGroup),
    names_to = c("Race", "Sex"),
    names_sep = "\\.",
    values_to = "Population"
  ) |>
  filter(Sex %in% c("Male", "Female")) |>
  add_race3_and_race2()

age_1970s_18_19 <- reshape_3cat(aggregate_3cat(Census_1970s_by_age |> filter(Age %in% c(18, 19)))) |>
  rename(Population_1819 = Population) |>
  add_race3_and_race2_1819()
```

The check below verifies that this era's by-age table reconciles with its
all-ages and adults totals, for both RaceScheme values. All four rows
should be exactly zero.

```{r check-age-group-1970s, dependson="age-group-1970s"}
bind_rows(
  AllAges = check_allages_reconciliation(
    census_targets_1970s_by_age, Census_1970s_allages, c("Year", "RaceScheme", "Race", "Sex")
  ),
  Adults = check_adults_reconciliation(
    census_targets_1970s_by_age, age_1970s_18_19, Census_1970s_adults, c("Year", "RaceScheme", "Race", "Sex")
  ),
  .id = "Check"
)
```

```{r age-group-1980s, dependson=c("load-RQI-files", "reshape-RQI", "bin-age-5yr", "check-allages-reconciliation", "check-adults-reconciliation")}
RQI_by_age <- RQI_raw |>
  mutate(AgeGroup = bin_age_5yr(Age)) |>
  group_by(SourceYear, AgeGroup) |>
  summarize(across(Total:API.NonHispanic.Female, sum), .groups = "drop") |>
  rename(Year = SourceYear)

census_targets_1980s_by_age <- RQI_by_age |>
  transmute(
    Year, AgeGroup,
    White.Male, White.Female,
    Black.Male, Black.Female,
    AIAN.Male = AIAE.Male, AIAN.Female = AIAE.Female,
    API.Male, API.Female
  ) |>
  pivot_longer(
    cols = -c(Year, AgeGroup),
    names_to = c("Race", "Sex"),
    names_sep = "\\.",
    values_to = "Population"
  ) |>
  add_race4_race3_race2()

census_targets_hispanic_by_age_1980s <- RQI_by_age |>
  transmute(
    Year, AgeGroup,
    White.Hispanic.Male = White.Male - White.NonHispanic.Male,
    White.Hispanic.Female = White.Female - White.NonHispanic.Female,
    White.NonHispanic.Male, White.NonHispanic.Female,
    Black.Hispanic.Male = Black.Male - Black.NonHispanic.Male,
    Black.Hispanic.Female = Black.Female - Black.NonHispanic.Female,
    Black.NonHispanic.Male, Black.NonHispanic.Female,
    AIAN.Hispanic.Male = AIAE.Male - AIAE.NonHispanic.Male,
    AIAN.Hispanic.Female = AIAE.Female - AIAE.NonHispanic.Female,
    AIAN.NonHispanic.Male = AIAE.NonHispanic.Male,
    AIAN.NonHispanic.Female = AIAE.NonHispanic.Female,
    API.Hispanic.Male = API.Male - API.NonHispanic.Male,
    API.Hispanic.Female = API.Female - API.NonHispanic.Female,
    API.NonHispanic.Male, API.NonHispanic.Female
  ) |>
  pivot_longer(
    cols = -c(Year, AgeGroup),
    names_to = c("Race", "HispanicOrigin", "Sex"),
    names_sep = "\\.",
    values_to = "Population"
  ) |>
  add_race4_race3_race2()

RQI_18_19 <- RQI_raw |>
  filter(Age %in% c(18, 19)) |>
  group_by(SourceYear) |>
  summarize(across(Total:API.NonHispanic.Female, sum), .groups = "drop") |>
  rename(Year = SourceYear)

age_1980s_18_19 <- reshape_RQI4(RQI_18_19) |>
  rename(Population_1819 = Population) |>
  add_race4_race3_race2_1819()
```

The check below verifies that this era's by-age table reconciles with its
all-ages and adults totals, for all three RaceScheme values. All six rows
should be exactly zero.

```{r check-age-group-1980s, dependson="age-group-1980s"}
bind_rows(
  AllAges = check_allages_reconciliation(
    census_targets_1980s_by_age, Census_1980s_allages, c("Year", "RaceScheme", "Race", "Sex")
  ),
  Adults = check_adults_reconciliation(
    census_targets_1980s_by_age, age_1980s_18_19, Census_1980s_adults, c("Year", "RaceScheme", "Race", "Sex")
  ),
  .id = "Check"
)
```

```{r age-group-1990s, dependson=c("derive-1990s-race-sex", "age-group-levels", "check-allages-reconciliation", "check-adults-reconciliation", "reshape-RQI")}
census_1990s_agegrp_labels <- c(
  `0` = "Under 5", `1` = "Under 5", `2` = "5 to 9", `3` = "10 to 14",
  `4` = "15 to 19", `5` = "20 to 24", `6` = "25 to 29", `7` = "30 to 34",
  `8` = "35 to 39", `9` = "40 to 44", `10` = "45 to 49", `11` = "50 to 54",
  `12` = "55 to 59", `13` = "60 to 64", `14` = "65 to 69", `15` = "70 to 74",
  `16` = "75 to 79", `17` = "80 to 84", `18` = "85 years and over"
)

census_targets_1990s_by_age <- census_1990s_race_sex_hisp_raw |>
  mutate(AgeGroup = census_1990s_agegrp_labels[as.character(AGEGRP)]) |>
  group_by(Year, Race, Sex, AgeGroup) |>
  summarize(Population = sum(POP), .groups = "drop") |>
  add_race4_race3_race2()

census_targets_hispanic_by_age_1990s <- census_1990s_race_sex_hisp_raw |>
  mutate(AgeGroup = census_1990s_agegrp_labels[as.character(AGEGRP)]) |>
  group_by(Year, Race, HispanicOrigin, Sex, AgeGroup) |>
  summarize(Population = sum(POP), .groups = "drop") |>
  add_race4_race3_race2()

# 1990-1999's own documented 2/5 "15 to 19" interpolation (see
# "1990-1999" above), applied identically here so this era's adults
# reconciliation is exact by construction, not an independent estimate.
# The interpolation is linear, so applying it after RaceScheme collapsing
# (as done here) gives the same result as applying it before.
age_1990s_18_19 <- census_targets_1990s_by_age |>
  filter(AgeGroup == "15 to 19") |>
  transmute(Year, RaceScheme, Race, Sex, Population_1819 = 0.4 * Population)
```

The check below verifies that this era's by-age table reconciles with its
all-ages and adults totals, for all three RaceScheme values. All six rows
should be exactly zero.

```{r check-age-group-1990s, dependson="age-group-1990s"}
bind_rows(
  AllAges = check_allages_reconciliation(
    census_targets_1990s_by_age, census_1990s_allages, c("Year", "RaceScheme", "Race", "Sex")
  ),
  Adults = check_adults_reconciliation(
    census_targets_1990s_by_age, age_1990s_18_19, census_1990s_adults, c("Year", "RaceScheme", "Race", "Sex")
  ),
  .id = "Check"
)
```

```{r age-group-2000-2019, dependson=c("combine-2000s-2010s", "age-group-levels", "check-allages-reconciliation", "check-adults-reconciliation")}
census_targets_2000_2019_by_age <- census_2000_2019_race_sex_raw |>
  filter(AGEGROUP %in% 1:18, HISP == 0) |>
  mutate(AgeGroup = age_group_levels[AGEGROUP]) |>
  group_by(Year, Race, Sex, AgeGroup) |>
  summarize(Population = sum(POP), .groups = "drop") |>
  add_race6_race3_race2()

census_targets_hispanic_by_age_2000_2019 <- census_2000_2019_race_sex_raw |>
  filter(AGEGROUP %in% 1:18, HISP %in% 1:2) |>
  mutate(
    AgeGroup = age_group_levels[AGEGROUP],
    HispanicOrigin = if_else(HISP == 2, "Hispanic", "NonHispanic")
  ) |>
  group_by(Year, Race, HispanicOrigin, Sex, AgeGroup) |>
  summarize(Population = sum(POP), .groups = "drop") |>
  add_race6_race3_race2()

# No single-year detail exists for this era (see "2000-2019" above), so
# this uses the same 2/5 "15 to 19" interpolation as 1990-1999 -- but
# unlike 1990-1999, this era's own adults table is built from an exact
# AGEGROUP = 29 code, not this interpolation, so the check below is a
# genuine (not circular) comparison and a small nonzero difference is
# expected, not a bug.
age_2000_2019_18_19 <- census_targets_2000_2019_by_age |>
  filter(AgeGroup == "15 to 19") |>
  transmute(Year, RaceScheme, Race, Sex, Population_1819 = 0.4 * Population)
```

The check below verifies that this era's by-age table reconciles with its
all-ages and adults totals, for all three RaceScheme values. The AllAges
rows should be exactly zero; the Adults rows will show a small nonzero
value, as explained above.

```{r check-age-group-2000-2019, dependson="age-group-2000-2019"}
bind_rows(
  AllAges = check_allages_reconciliation(
    census_targets_2000_2019_by_age, census_2000_2019_allages, c("Year", "RaceScheme", "Race", "Sex")
  ),
  Adults = check_adults_reconciliation(
    census_targets_2000_2019_by_age, age_2000_2019_18_19, census_2000_2019_adults, c("Year", "RaceScheme", "Race", "Sex")
  ),
  .id = "Check"
)
```

```{r age-group-2020s, dependson=c("load-2020s-census-files", "summarize-2020s-race-sex", "bin-age-5yr", "check-allages-reconciliation", "check-adults-reconciliation", "combine-2000s-2010s")}
census_2020s_by_age_raw <- census_2020s_raw |>
  filter(AGE != 999) |>
  mutate(AgeGroup = bin_age_5yr(AGE))

census_targets_2020s_by_age <- census_2020s_by_age_raw |>
  group_by(YEAR, AgeGroup) |>
  summarize(
    White_Male = sum(WA_MALE), White_Female = sum(WA_FEMALE),
    Black_Male = sum(BA_MALE), Black_Female = sum(BA_FEMALE),
    AIAN_Male = sum(IA_MALE), AIAN_Female = sum(IA_FEMALE),
    Asian_Male = sum(AA_MALE), Asian_Female = sum(AA_FEMALE),
    NHPI_Male = sum(NA_MALE), NHPI_Female = sum(NA_FEMALE),
    TwoOrMore_Male = sum(TOM_MALE), TwoOrMore_Female = sum(TOM_FEMALE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = -c(YEAR, AgeGroup),
    names_to = c("Race", "Sex"),
    names_sep = "_",
    values_to = "Population"
  ) |>
  rename(Year = YEAR) |>
  add_race6_race3_race2()

census_targets_hispanic_by_age_2020s <- census_2020s_by_age_raw |>
  group_by(YEAR, AgeGroup) |>
  summarize(
    White.Hispanic_Male = sum(HWA_MALE), White.Hispanic_Female = sum(HWA_FEMALE),
    White.NonHispanic_Male = sum(NHWA_MALE), White.NonHispanic_Female = sum(NHWA_FEMALE),
    Black.Hispanic_Male = sum(HBA_MALE), Black.Hispanic_Female = sum(HBA_FEMALE),
    Black.NonHispanic_Male = sum(NHBA_MALE), Black.NonHispanic_Female = sum(NHBA_FEMALE),
    AIAN.Hispanic_Male = sum(HIA_MALE), AIAN.Hispanic_Female = sum(HIA_FEMALE),
    AIAN.NonHispanic_Male = sum(NHIA_MALE), AIAN.NonHispanic_Female = sum(NHIA_FEMALE),
    Asian.Hispanic_Male = sum(HAA_MALE), Asian.Hispanic_Female = sum(HAA_FEMALE),
    Asian.NonHispanic_Male = sum(NHAA_MALE), Asian.NonHispanic_Female = sum(NHAA_FEMALE),
    NHPI.Hispanic_Male = sum(HNA_MALE), NHPI.Hispanic_Female = sum(HNA_FEMALE),
    NHPI.NonHispanic_Male = sum(NHNA_MALE), NHPI.NonHispanic_Female = sum(NHNA_FEMALE),
    TwoOrMore.Hispanic_Male = sum(HTOM_MALE), TwoOrMore.Hispanic_Female = sum(HTOM_FEMALE),
    TwoOrMore.NonHispanic_Male = sum(NHTOM_MALE), TwoOrMore.NonHispanic_Female = sum(NHTOM_FEMALE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = -c(YEAR, AgeGroup),
    names_to = c("Race", "HispanicOrigin", "Sex"),
    names_sep = "[._]",
    values_to = "Population"
  ) |>
  rename(Year = YEAR) |>
  add_race6_race3_race2()

census_2020s_1819 <- census_2020s_by_age_raw |>
  filter(AGE %in% c(18, 19)) |>
  summarize_2020s_race_sex() |>
  rename(Population_1819 = Population) |>
  add_race6_race3_race2_1819()
```

The check below verifies that this era's by-age table reconciles with its
all-ages and adults totals, for all three RaceScheme values. All six rows
should be exactly zero.

```{r check-age-group-2020s, dependson="age-group-2020s"}
bind_rows(
  AllAges = check_allages_reconciliation(
    census_targets_2020s_by_age, census_2020s_race_sex_allages, c("Year", "RaceScheme", "Race", "Sex")
  ),
  Adults = check_adults_reconciliation(
    census_targets_2020s_by_age, census_2020s_1819, census_2020s_race_sex_adults, c("Year", "RaceScheme", "Race", "Sex")
  ),
  .id = "Check"
)
```

```{r combine-age-group-1960-2024, dependson=c("age-group-1960s", "age-group-1970s", "age-group-1980s", "age-group-1990s", "age-group-2000-2019", "age-group-2020s", "age-group-levels")}
census_targets_by_age <- bind_rows(
  census_targets_1960s_by_age, census_targets_1970s_by_age, census_targets_1980s_by_age,
  census_targets_1990s_by_age, census_targets_2000_2019_by_age, census_targets_2020s_by_age
) |>
  mutate(AgeGroup = factor(AgeGroup, levels = age_group_levels)) |>
  arrange(Year, RaceScheme, Race, Sex, AgeGroup) |>
  mutate(AgeGroup = as.character(AgeGroup))
```

```{r combine-age-group-hispanic, dependson=c("age-group-1980s", "age-group-1990s", "age-group-2000-2019", "age-group-2020s", "age-group-levels")}
census_targets_hispanic_by_age <- bind_rows(
  census_targets_hispanic_by_age_1980s, census_targets_hispanic_by_age_1990s,
  census_targets_hispanic_by_age_2000_2019, census_targets_hispanic_by_age_2020s
) |>
  mutate(AgeGroup = factor(AgeGroup, levels = age_group_levels)) |>
  arrange(Year, RaceScheme, Race, HispanicOrigin, Sex, AgeGroup) |>
  mutate(AgeGroup = as.character(AgeGroup))
```

## Final Deliverables: Two Consolidated Files

Both files share the same seven columns, `Year, RaceScheme, Race, Sex,
AgeGroup, HispanicOrigin, Population`, spanning 1900-2025. File 1 covers
all ages; File 2 exists only because Census's own "15 to 19" bin straddles
the 18+ adult/minor line, so an all-ages file alone can't recover a clean
18+ total by summing bins -- File 2 resolves that with a residual `"18 to
19 years"` bin instead.

```{r build-final-by-age, dependson=c("combine-1900-1959", "combine-all-eras", "age-group-1900-1959", "combine-age-group-1960-2024", "combine-age-group-hispanic")}
census_targets_final_by_age <- bind_rows(
  census_targets_1900_1959_by_age |> mutate(HispanicOrigin = NA_character_),
  census_targets_by_age |> filter(Year < 1980) |> mutate(HispanicOrigin = NA_character_),
  census_targets_hispanic_by_age
) |>
  select(Year, RaceScheme, Race, Sex, AgeGroup, HispanicOrigin, Population) |>
  arrange(Year, RaceScheme, Race, Sex, HispanicOrigin, AgeGroup)

write_csv(
  census_targets_final_by_age,
  file.path(artifacts_dir, "census_targets.csv")
)
```

File 1 (all ages) has now been written. The row and year counts below
confirm it spans the full 1900-2025 range.

```{r report-final-by-age, dependson="build-final-by-age"}
cat(
  "File 1 (all ages): rows =", nrow(census_targets_final_by_age),
  "| years =", min(census_targets_final_by_age$Year), "-", max(census_targets_final_by_age$Year)
)
```

The check below verifies that, for every RaceScheme and every 1980+ cell,
Hispanic plus NonHispanic sums exactly to that cell's Race total, computed
independently. This should be exactly zero.

```{r check-final-by-age-no-doublecount, dependson=c("build-final-by-age", "combine-age-group-1960-2024")}
census_targets_by_age |>
  filter(Year >= 1980) |>
  rename(Population_RaceTotal = Population) |>
  inner_join(
    census_targets_hispanic_by_age |>
      group_by(Year, RaceScheme, Race, Sex, AgeGroup) |>
      summarize(Population_HispanicSum = sum(Population), .groups = "drop"),
    by = c("Year", "RaceScheme", "Race", "Sex", "AgeGroup")
  ) |>
  mutate(Diff = Population_HispanicSum - Population_RaceTotal) |>
  summarize(RowsChecked = n(), MaxAbsDiff = max(abs(Diff)))
```

The check below verifies that File 1, summed across AgeGroup and
HispanicOrigin, reproduces the independently tabulated all-ages totals for
every RaceScheme. This should be exactly zero.

```{r check-final-by-age-vs-allages, dependson=c("build-final-by-age", "combine-1900-1959", "combine-all-eras")}
census_targets_final_by_age |>
  group_by(Year, RaceScheme, Race, Sex) |>
  summarize(Population_FromFile1 = sum(Population), .groups = "drop") |>
  inner_join(
    bind_rows(census_targets_1900_1959_allages, census_targets_allages),
    by = c("Year", "RaceScheme", "Race", "Sex")
  ) |>
  mutate(Diff = Population_FromFile1 - Population) |>
  summarize(RowsChecked = n(), MaxAbsDiff = max(abs(Diff)))
```

The check below verifies that every RaceScheme available for a given
Year/Sex/AgeGroup/HispanicOrigin cell sums to the same total population.
Coarser schemes are always derived by collapsing a finer one (see
"Overview" above), so this should be exactly zero; a nonzero value here
would indicate a mistake in one of the collapse maps.

```{r check-final-by-age-cross-scheme, dependson="build-final-by-age"}
census_targets_final_by_age |>
  group_by(Year, RaceScheme, Sex, AgeGroup, HispanicOrigin) |>
  summarize(Population = sum(Population), .groups = "drop") |>
  group_by(Year, Sex, AgeGroup, HispanicOrigin) |>
  summarize(MaxAbsDiffAcrossSchemes = max(Population) - min(Population), .groups = "drop") |>
  summarize(MaxAbsDiff = max(MaxAbsDiffAcrossSchemes))
```

```{r build-final-adults-by-age, dependson=c("build-final-by-age", "combine-1900-1959", "combine-all-eras", "combine-hispanic", "age-group-levels")}
# "Full" bins are everything strictly above "15 to 19" (an exclusion, not
# an inclusion list, since 1900-1959's own open-ended "75 years and over"
# catch-all isn't one of the standard 18 age_group_levels bins but is
# entirely 18+).
young_bins <- age_group_levels[1:4]

# Splits a by-age table into its full (unambiguously 18+) bins, and a
# residual "18 to 19 years" bin computed as adults_total - sum(full bins).
build_adults_by_age <- function(by_age, adults_total, group_vars) {
  full_bins <- by_age |>
    filter(!AgeGroup %in% young_bins) |>
    select(all_of(group_vars), AgeGroup, Population)

  full_bin_sums <- full_bins |>
    group_by(across(all_of(group_vars))) |>
    summarize(Population_FullBins = sum(Population), .groups = "drop")

  residual_1819 <- adults_total |>
    inner_join(full_bin_sums, by = group_vars) |>
    mutate(
      AgeGroup = "18 to 19 years",
      Population = Population - Population_FullBins
    ) |>
    select(all_of(group_vars), AgeGroup, Population)

  bind_rows(full_bins, residual_1819)
}

adults_by_age_1900_1959 <- build_adults_by_age(
  census_targets_1900_1959_by_age, census_targets_1900_1959_adults, c("Year", "RaceScheme", "Race", "Sex")
) |>
  mutate(HispanicOrigin = NA_character_)

adults_by_age_1960_1979 <- build_adults_by_age(
  census_targets_by_age |> filter(Year < 1980),
  census_targets_adults |> filter(Year < 1980),
  c("Year", "RaceScheme", "Race", "Sex")
) |>
  mutate(HispanicOrigin = NA_character_)

adults_by_age_1980_2024 <- build_adults_by_age(
  census_targets_hispanic_by_age, census_targets_hispanic_adults,
  c("Year", "RaceScheme", "Race", "Sex", "HispanicOrigin")
)

census_targets_final_adults_by_age <- bind_rows(
  adults_by_age_1900_1959, adults_by_age_1960_1979, adults_by_age_1980_2024
) |>
  select(Year, RaceScheme, Race, Sex, AgeGroup, HispanicOrigin, Population) |>
  arrange(Year, RaceScheme, Race, Sex, HispanicOrigin, AgeGroup)

write_csv(
  census_targets_final_adults_by_age,
  file.path(artifacts_dir, "census_targets_adults.csv")
)
```

File 2 (adults 18+) has now been written. The row and year counts below
confirm it spans the full 1900-2025 range.

```{r report-final-adults-by-age, dependson="build-final-adults-by-age"}
cat(
  "File 2 (adults 18+): rows =", nrow(census_targets_final_adults_by_age),
  "| years =", min(census_targets_final_adults_by_age$Year), "-", max(census_targets_final_adults_by_age$Year)
)
```

The check below verifies that File 2 reproduces the existing adults
totals, with and without the HispanicOrigin split, for every RaceScheme.
Both rows should be exactly zero.

```{r check-final-adults-by-age, dependson=c("build-final-adults-by-age", "combine-1900-1959", "combine-all-eras", "combine-hispanic")}
check_totals <- census_targets_final_adults_by_age |>
  group_by(Year, RaceScheme, Race, Sex) |>
  summarize(Population_FromFile2 = sum(Population), .groups = "drop") |>
  inner_join(
    bind_rows(census_targets_1900_1959_adults, census_targets_adults),
    by = c("Year", "RaceScheme", "Race", "Sex")
  ) |>
  mutate(Diff = Population_FromFile2 - Population)

check_hispanic <- census_targets_final_adults_by_age |>
  filter(!is.na(HispanicOrigin)) |>
  group_by(Year, RaceScheme, Race, Sex, HispanicOrigin) |>
  summarize(Population_FromFile2 = sum(Population), .groups = "drop") |>
  inner_join(census_targets_hispanic_adults, by = c("Year", "RaceScheme", "Race", "Sex", "HispanicOrigin")) |>
  mutate(Diff = Population_FromFile2 - Population)

bind_rows(
  `vs. Year/RaceScheme/Race/Sex adults totals` = summarize(check_totals, RowsChecked = n(), MaxAbsDiff = max(abs(Diff))),
  `vs. Year/RaceScheme/Race/Sex/HispanicOrigin adults totals (1980+)` = summarize(check_hispanic, RowsChecked = n(), MaxAbsDiff = max(abs(Diff))),
  .id = "Check"
)
```

The check below verifies that every RaceScheme available for a given
Year/Sex/AgeGroup/HispanicOrigin cell sums to the same total population,
mirroring the File 1 cross-scheme check above, and that the residual "18
to 19 years" bin is never negative.

```{r check-final-adults-by-age-cross-scheme, dependson="build-final-adults-by-age"}
census_targets_final_adults_by_age |>
  group_by(Year, RaceScheme, Sex, AgeGroup, HispanicOrigin) |>
  summarize(Population = sum(Population), .groups = "drop") |>
  group_by(Year, Sex, AgeGroup, HispanicOrigin) |>
  summarize(MaxAbsDiffAcrossSchemes = max(Population) - min(Population), .groups = "drop") |>
  summarize(MaxAbsDiff = max(MaxAbsDiffAcrossSchemes))

census_targets_final_adults_by_age |>
  filter(AgeGroup == "18 to 19 years") |>
  summarize(MinPopulation_1819 = min(Population))
```
