---
title: "Census PEP Population Targets: Wide-Format Workbooks"
author: "Joshua Born"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
library(dplyr)
library(here)
library(readr)
library(tidyr)
library(writexl)

artifacts_dir <- here("output")
dir.create(artifacts_dir, showWarnings = FALSE, recursive = TRUE)

knitr::opts_chunk$set(
  echo = TRUE,
  tidy = TRUE
)
```

## Overview

Builds four wide-format `.xlsx` workbooks for spreadsheet users, from the
two long/tidy CSVs this project publishes
(`census_targets.csv`, `census_targets_adults.csv`) -- the same files
described in the README's "Picking a File" section, consumed here exactly
as an external user would consume them, not any in-memory objects from
`census_targets.Rmd`.

Each workbook has one sheet per `RaceScheme`, with one row per `Year` and
one column per remaining combination of variables. Two workbooks
(`census_targets_wide.xlsx`, `census_targets_wide_adults.xlsx`) cover all
four race schemes (`race2`, `race3`, `race4`, `race6`) but sum away the
`HispanicOrigin` split, since that split only exists from 1980 on and
therefore isn't available for every sheet's full year range. The other two
workbooks (`census_targets_wide_hispanic.xlsx`,
`census_targets_wide_adults_hispanic.xlsx`) keep the `HispanicOrigin`
split, and so are limited to the `race4` (1980-1999) and `race6`
(2000-2024) sheets, which together cover the entire 1980-2024 span over
which `HispanicOrigin` is available.

```{r read-inputs}
census_targets <- read_csv(file.path(artifacts_dir, "census_targets.csv"), show_col_types = FALSE)
census_targets_adults <- read_csv(file.path(artifacts_dir, "census_targets_adults.csv"), show_col_types = FALSE)
```

## Main Wide Workbooks (No Hispanic-Origin Split)

For each `RaceScheme`, this collapses `HispanicOrigin` by summing it away,
then pivots so that each row is a `Year` and each column is a
`Race_Sex_AgeGroup` combination.

```{r wide-no-hispanic-function}
race_schemes <- c("race2", "race3", "race4", "race6")

build_wide_sheets <- function(long_data) {
  sheets <- lapply(race_schemes, function(scheme) {
    long_data |>
      filter(RaceScheme == scheme) |>
      group_by(Year, Race, Sex, AgeGroup) |>
      summarize(Population = sum(Population), .groups = "drop") |>
      mutate(Cell = paste(Race, Sex, AgeGroup, sep = "_")) |>
      select(Year, Cell, Population) |>
      pivot_wider(id_cols = Year, names_from = Cell, values_from = Population) |>
      arrange(Year)
  })
  names(sheets) <- race_schemes
  sheets
}

wide_sheets <- build_wide_sheets(census_targets)
wide_adults_sheets <- build_wide_sheets(census_targets_adults)

write_xlsx(wide_sheets, file.path(artifacts_dir, "census_targets_wide.xlsx"))
write_xlsx(wide_adults_sheets, file.path(artifacts_dir, "census_targets_wide_adults.xlsx"))
```

`census_targets_wide.xlsx` and `census_targets_wide_adults.xlsx` have now
been written, each with sheets `r paste(race_schemes, collapse = ", ")`.

The check below verifies that, for every sheet, the total population in
the wide sheet matches the total population in the corresponding filtered
long data. All values should be exactly zero.

```{r check-wide-no-hispanic, dependson=c("wide-no-hispanic-function", "read-inputs")}
check_wide_sums <- function(long_data, wide_sheets) {
  bind_rows(lapply(race_schemes, function(scheme) {
    long_total <- long_data |>
      filter(RaceScheme == scheme) |>
      summarize(Population = sum(Population)) |>
      pull(Population)

    wide_total <- wide_sheets[[scheme]] |>
      select(-Year) |>
      summarize(across(everything(), \(x) sum(x, na.rm = TRUE))) |>
      rowSums()

    tibble(RaceScheme = scheme, Diff = wide_total - long_total)
  }))
}

bind_rows(
  AllAges = check_wide_sums(census_targets, wide_sheets),
  Adults = check_wide_sums(census_targets_adults, wide_adults_sheets),
  .id = "File"
)
```

## Hispanic-Origin Wide Workbooks

For `race4` (1980-1999) and `race6` (2000-2024), this pivots without
summing away `HispanicOrigin`, so each column is a
`Race_Sex_AgeGroup_HispanicOrigin` combination.

```{r wide-hispanic-function}
hispanic_schemes <- c("race4", "race6")

build_wide_hispanic_sheets <- function(long_data) {
  sheets <- lapply(hispanic_schemes, function(scheme) {
    long_data |>
      filter(RaceScheme == scheme, !is.na(HispanicOrigin)) |>
      mutate(Cell = paste(Race, Sex, AgeGroup, HispanicOrigin, sep = "_")) |>
      select(Year, Cell, Population) |>
      pivot_wider(id_cols = Year, names_from = Cell, values_from = Population) |>
      arrange(Year)
  })
  names(sheets) <- hispanic_schemes
  sheets
}

wide_hispanic_sheets <- build_wide_hispanic_sheets(census_targets)
wide_adults_hispanic_sheets <- build_wide_hispanic_sheets(census_targets_adults)

write_xlsx(wide_hispanic_sheets, file.path(artifacts_dir, "census_targets_wide_hispanic.xlsx"))
write_xlsx(wide_adults_hispanic_sheets, file.path(artifacts_dir, "census_targets_wide_adults_hispanic.xlsx"))
```

`census_targets_wide_hispanic.xlsx` and
`census_targets_wide_adults_hispanic.xlsx` have now been written, each with
sheets `r paste(hispanic_schemes, collapse = ", ")`.

The check below verifies that, for every sheet, the total population in
the wide sheet matches the total population in the corresponding filtered
long data (restricted to rows with a non-missing `HispanicOrigin`). All
values should be exactly zero.

```{r check-wide-hispanic, dependson=c("wide-hispanic-function", "read-inputs")}
check_wide_hispanic_sums <- function(long_data, wide_sheets) {
  bind_rows(lapply(hispanic_schemes, function(scheme) {
    long_total <- long_data |>
      filter(RaceScheme == scheme, !is.na(HispanicOrigin)) |>
      summarize(Population = sum(Population)) |>
      pull(Population)

    wide_total <- wide_sheets[[scheme]] |>
      select(-Year) |>
      summarize(across(everything(), \(x) sum(x, na.rm = TRUE))) |>
      rowSums()

    tibble(RaceScheme = scheme, Diff = wide_total - long_total)
  }))
}

bind_rows(
  AllAges = check_wide_hispanic_sums(census_targets, wide_hispanic_sheets),
  Adults = check_wide_hispanic_sums(census_targets_adults, wide_adults_hispanic_sheets),
  .id = "File"
)
```
